The benefits of reporting critical effect size values

### 2024-08-20 paper/paper.Rmd
m

abstract: |
  Critical effect size values represent the smallest detectable effect that would reach statistical significance given a certain sample size. These values are highly applicable in various scenarios, as they do not require specifying a plausible effect size of interest. Reporting critical effect size values may be especially useful when adopting NHST (Null Hypothesis Significance Testing) and the sample size has not been planned a priori, or when there is uncertainty on the expected effect size. To assist researchers in calculating critical effect size values, we have developed an R package that allows to report the critical effect size values of model parameters. The package's functions are implemented for group comparisons, correlations, linear regression, and meta-analysis. What we propose could benefit researchers during the planning phase of the study by helping them to understand the limitations and strengths of their research design, and also as a retrospective tool to potentially reframe original interpretations.

keywords          : critical effect size values, statistical significance, fucntions implementation 
wordcount         : "X"

@@

In the present paper, we suggest that researchers systematically report the critical effect size values when they make inference using statistical significance. The "critical effect size values" refers to the smallest statistically significant effect sizes, which depends on the test performed and the characteristics of the sample [@lakens2022sample]. To cut to the chase, let us consider a study that tests (using for example a two-tailed test) a bivariate correlation with $n = 20$ participants and uses $p < 0.05$ as the threshold for statistical significance: in this case, the smallest significant effects (critical effect size values) are Pearson's $r = -0.44$ and $r = 0.44$. As we will argue, the critical effect size values can serve as a useful complement to aid in the interpretation of study results, communicate their relevance, help assessing their replicability, particularly when statistical significance is relied upon for inference. This is especially important in cases where the sample size was not (or could not) be predetermined based on a theory-driven target effect size or when statistical power may be difficult to determine or deviates from an optimal level.

Despite longstanding criticisms, the Null Hypothesis Significance Testing (NHST) remains a prominent approach for statistical inference in science [@cohen1994; @gigerenzer2004null; @krueger2001null]. Over the past decades, numerous proposals have emerged to enhance inference based on this approach. These include conducting power analyses to determine sample size ($n$) a priori based on theory-informed target effects, and reporting of effect sizes and their confidence intervals, accompanied by pertinent related comments, rather than drawing dichotomous conclusions solely based on statistical significance [@transue2019apa]. Even for the most skeptical, there are suggestions that NHST may possess merits of its own. For instance, @wilson2020science propose that it could serve as a useful preliminary filter utilized by "original science" for screening potentially interesting effects, that should subsequently be validated by "replication science".

In psychology, power is often severely limited for medium to small effect sizes [@szucs2017empirical], which are generally expected to represent true replicable findings. In fact, resource constraints have been authoritatively listed among acceptable justifications for sample size  [@lakens2022sample], and this scenario is arguably widespread in psychology. However, limited power makes it challenging to distinguish signal from noise, and on average it results in an overestimation of effect sizes when relying on statistical significance for inference [@altoe2020enhancing]. On the other hand, very large samples are occasionally available in psychological science, albeit less frequently. However, as some have argued, excessively large sample sizes risk creating an "everything-is-significant" scenario, where researchers report and discuss practically negligible effects solely because they are statistically significant, regardless of theoretical relevance or whether they just reflect minor procedural artifacts [@wilson2020science]. Thus, while statistical significance can serve as a potentially useful filter, relying solely on it for inference may introduce risks in interpretation. In such cases, we contend that reporting critical effect size values can help put into perspective and convey the relevance of the results.

There are two clear benefits of reporting the critical effect size value for a corresponding test. First, when sample sizes are small, the critical effect size values inform readers about whether the effect sizes that could lead to a rejection of the null hypothesis are in line with realistic expectations. If the sample size is small, and only very large effects would yield a statistically significant result, and the underlying mechanism that is examined is unlikely to lead to such large effect sizes, in these cases, researchers will realized they are not able to collect sufficient data to perform a meaningful test. An a-priori power analysis would typically lead to a similar conclusion, but reporting the critical effect size values focuses the attention more strongly on which effect sizes are reasonable to expect. Second, in large studies the critical effect size values will make clear that trivially small effect sizes will be statistically significant. This will focus the attention of researchers on the difference between statistical significance and practical significance. This is especially important in correlational studies on large samples in psychology, where small uncontrolled sources of statistical noise may simultaneously affect and create some shared variance across otherwise unrelated variables, a phenomenon controversially labelled as the 'crud factor' [@orben_crud_2020].

Let us consider the following two scenarios as examples. First, imagine researchers conducting a study involving a between-group comparison. Due to severe resource constraints, they are only able to collect a limited sample size of $n = 30$ ($n = 15$ per group). No statistically significant effect is detected with $p < 0.05$. Given their prior expectation that the effect of interest may not be large, they acknowledge that their study was likely underpowered, though uncertain to some extent. Subsequently, they compute the critical effect size values, revealing Cohen's $d = -0.75$ and $d = 0.75$. This indicates that any estimated |Cohen's $d| < 0.75$ will certainly fail to reach statistical significance. By reporting these critical effect size values, the researchers transparently convey that estimated effects up to a medium-to-large magnitude will consistently fall short of significance. In essence, this scenario exemplifies a version of the "winner's curse" [@hedges1984estimation], made explicit. In this context, the winner's curse indicates the tendency for a statistically significant initial finding (i.e., a "winner") to be associated with an overestimated effect size which will probably be deflated in subsequent replications (i.e., the "curse"), especially if such initial finding was obtained on a small sample. Knowing from the critical effect size values that a specific result must necessarily be associated with an estimated large effect size or be non-significant, readers are, in principle, prevented from both over-interpreting the magnitude of an effect size and from equating lack of statistical significance with lack of an effect of interest.

In a second scenario, imagine researchers who gain access to a very large archival dataset ($n = 5,000$) and decide to explore some bivariate correlations. With such a large sample size, researchers are not constrained by sample size limitations and could instead just focus on effect sizes. However, they opt to legitimately use statistical significance of a two-tailed test as their primary criterion for whether to comment on effects in their further discussions. In this context, the critical effect size values for significance are determined to be correlations of $r = \pm 0.03$. This implies that practically any effect differing from zero, regardless of its small magnitude, may easily attain statistical significance. It is worth noting that such small effects may potentially reflect just minor artifacts [@wilson2020science], such as subtle experimenter effects or slight non-independence among observations, even in meticulously designed studies. While readers may be tempted to interpret any reported effects simply because they attain statistical significance, preliminarily signaling the critical effect size value level can serve as a clear warning that the significance filter might be somehow permissive, thus urging interpretive caution.

A similar scenario may arise in meta-analysis. Despite potential loss of precision due to substantial heterogeneity across effect sizes in different studies, meta-analyses typically synthesize a large amount of evidence, leading even very small average effect sizes to reach significance. While the focus of meta-analysis is generally on estimating effect sizes with uncertainty, statistical significance is routinely reported and interpreted. Thus, in meta-analysis as well, reporting the critical effect size values can caution readers that the mere attainment of statistical significance by the average effect does not automatically ensure scientific or real-life relevance.

To provide more tangible illustrations of the practical application of critical effect size values, let us examine two real-world instances drawn from published research. The first one features a small sample, and has been targeted in the series of replications by the Open Science Collaboration [@open2015estimating], while the second one presents with a very large sample and has already attracted attention in the scientific community. The first case pertains to "Study 5" conducted by @mccrea2008self. This experiment compared two groups based on their performance percentage in a math test following a preliminary practice session with feedback of failure. One group was exposed to failure-excusing, self-handicapping thoughts, while both groups underwent the math test afterward. With a modest sample size of 28 participants split into two groups (13 and 15 participants, respectively), a directional one-tailed t-test was conducted on the percentage of correct answers to compare the groups. The resulting effect was statistically significant, $t(26) = 1.87, p (one-tailed) < .05$. Based on the reported data, Cohen's $d = 0.736$, indicating a considerable effect size. However, the critical value for Cohen's $d$ in this context was $0.645$, suggesting that, given this sample size, statistical significance is achieved only for estimated effects of substantial magnitude. Subsequent replication efforts [@open2015estimating] reaffirmed the significance of the effect, and used a somewhat larger sample ($n = 61$), $t(59)=2.325$. The estimated effect size was somehow lower than that originally reported, however, $d = 0.605$ (here, critical d = 0.428). The second scenario involves the study by @kramer2014experimental, which explored the impact of emotional content on Facebook users' experiences. With a notably large sample size of $n = 689'003$, the researchers observed a significant effect worsened emotional states when positive posts were reduced, $t(310,044) = −5.63$, $p < 0.001$ (non-directional test). Based on this and other results, the authors conclude that emotions expressed by other users on Facebook influence our own emotions. Despite the statistical significance, however, the critical values for Cohen's $d$ in this case are $-0.006$ and $+0.006$, emphasizing the importance of considering effect size beyond mere significance. This prompts reflection on whether an effect of such magnitude on such a large sample (actual $d = 0.02$) is truly meaningful in individuals' lives.

>>>>

abstract: |
  Critical effect size values represent the smallest detectable effect that would reach statistical significance given a certain sample size. These values are highly applicable in various scenarios, as they do not require specifying a plausible effect size of interest. Reporting critical effect size values may be especially useful when adopting NHST (Null Hypothesis Significance Testing) and the sample size has not been planned a priori, or when there is uncertainty about the expected effect size. To assist researchers in calculating critical effect size values, we have developed an R package that allows to report the critical effect size values of model parameters. The package's functions are implemented for group comparisons, correlations, linear regression, and meta-analysis. What we propose could benefit researchers during the planning phase of the study by helping them to understand the limitations and strengths of their research design, and also as a retrospective tool to potentially reframe original interpretations.

keywords          : critical effect size values, statistical significance, fucntions implementation 
wordcount         : "3900"

@@

In the present paper, we suggest that researchers should systematically report critical effect size values when they make inference using statistical significance. The "critical effect size values" refers to the smallest statistically significant effect sizes, which depend on the test performed and the characteristics of the sample [@lakens2022sample]. To cut to the chase, let us consider a study that tests (using for example a two-tailed test) a bivariate correlation with $n = 20$ participants and uses $p < 0.05$ as the threshold for statistical significance: in this case, the smallest significant effects (critical effect size values) are Pearson's $r = -0.44$ and $r = 0.44$. As we will argue, critical effect size values can serve as a useful complement to aid in the interpretation of study results, to communicate their relevance, and to help assess their replicability, particularly when statistical significance is relied upon for inference. This is especially important in cases where the sample size was not (or could not) be predetermined based on a theory-driven target effect size, or when statistical power may be difficult to determine or deviates from an optimal level.

Despite longstanding criticisms, Null Hypothesis Significance Testing (NHST) remains a prominent approach for statistical inference in science [@cohen1994; @gigerenzer2004null; @krueger2001null]. Over the past decades, numerous proposals have emerged to enhance inference based on this approach. These include conducting power analyses to determine sample size ($n$) a priori based on theory-informed target effects, and reporting effect sizes and their confidence intervals, accompanied by pertinent related comments, rather than drawing dichotomous conclusions solely based on statistical significance [@transue2019apa]. Even for the most skeptical, there are suggestions that NHST may possess merits of its own. For instance, @wilson2020science propose that it could serve as a useful preliminary filter utilized by "original science" for screening potentially interesting effects, which should subsequently be validated by "replication science".

In psychology, statistical power is often severely limited for medium to small effect sizes [@szucs2017empirical], which are generally expected to represent true replicable findings. In fact, resource constraints have been authoritatively listed among acceptable justifications for sample size  [@lakens2022sample], and this scenario is arguably widespread in psychology. However, limited power makes it challenging to distinguish signal from noise, and on average it results in an overestimation of effect sizes when relying on statistical significance for inference [@altoe2020enhancing]. On the other hand, very large samples are occasionally available in psychological science, albeit less frequently. However, as some have argued, excessively large sample sizes risk creating an "everything-is-significant" scenario, where researchers report and discuss practically negligible effects solely because they are statistically significant, regardless of theoretical relevance or whether they merely reflect minor procedural artifacts [@wilson2020science]. Thus, while statistical significance can serve as a potentially useful filter, relying solely on it for inference may introduce risks in interpretation. In such cases, we contend that reporting critical effect size values can help put into perspective and convey the relevance of the results.

There are two clear benefits to reporting the critical effect size value for a corresponding test. First, when sample sizes are small, the critical effect size values inform readers about whether the effect sizes that could lead to rejecting the null hypothesis are in line with realistic expectations. If the sample size is small and only very large effects would yield a statistically significant result, and the underlying mechanism that is examined is unlikely to lead to such large effect sizes, in these cases, researchers will realize they are not able to collect sufficient data to perform a meaningful test. An a-priori power analysis would typically lead to a similar conclusion, but reporting the critical effect size values focuses the attention more strongly on which effect sizes are reasonable to expect. Second, in large studies the critical effect size values will make it clear that trivially small effect sizes will be statistically significant. This will focus the attention of researchers on the difference between statistical significance and practical significance. This is especially important in correlational studies on large samples in psychology, where small uncontrolled sources of statistical noise may simultaneously affect and create some shared variance across otherwise unrelated variables, a phenomenon controversially labelled as the 'crud factor' [@orben_crud_2020].

Let us consider the following two scenarios as examples. First, imagine researchers conducting a study involving a between-group comparison. Due to severe resource constraints, they are only able to collect a limited sample size of $n = 30$ ($n = 15$ per group). No statistically significant effect is detected with $p < 0.05$. Given their prior expectation that the effect of interest may not be large, they acknowledge that their study was likely underpowered, though uncertain to some extent. Subsequently, they compute the critical effect size values, revealing Cohen's $d = -0.75$ and $d = 0.75$. This indicates that any estimated |Cohen's $d| < 0.75$ will certainly fail to reach statistical significance. By reporting these critical effect size values, the researchers transparently convey that estimated effects up to a medium-to-large magnitude will consistently fall short of significance. In essence, this scenario exemplifies a version of the "winner's curse" [@hedges1984estimation], made explicit. In this context, the winner's curse indicates the tendency for a statistically significant initial finding (i.e., a "winner") to be associated with an overestimated effect size which will likely be deflated in subsequent replications (i.e., the "curse"), especially if such initial finding was obtained with a small sample. Knowing from the critical effect size values that a specific result must necessarily be associated with either an estimated large effect size or be non-significant, readers are, in principle, prevented from both over-interpreting the magnitude of an effect size and from equating lack of statistical significance with lack of an effect of interest.

In a second scenario, imagine researchers who gain access to a very large archival dataset ($n = 5,000$) and decide to explore some bivariate correlations. With such a large sample size, researchers are not constrained by sample size limitations and could instead focus on effect sizes. However, they legitimately  opt to use statistical significance of a two-tailed test as their primary criterion for whether to comment on effects in their further discussions. In this context, the critical effect size values for significance are determined to be correlations of $r = \pm 0.03$. This implies that practically any effect differing from zero, regardless of its small magnitude, may easily attain statistical significance. It is worth noting that such small effects may potentially reflect just minor artifacts [@wilson2020science], such as subtle experimenter effects or slight non-independence among observations, even in meticulously designed studies. While readers may be tempted to interpret any reported effects simply because they attain statistical significance, signaling the critical effect size value level beforehand can serve as a clear warning that the significance filter might be overly permissive, thus urging interpretive caution.

A similar scenario may arise in meta-analysis. Despite potential loss of precision due to substantial heterogeneity across effect sizes in different studies, meta-analyses typically synthesize a large amount of evidence, leading even very small average effect sizes to reach significance. While the focus of meta-analysis is generally on estimating effect sizes with uncertainty, statistical significance is routinely reported and interpreted. Thus, in meta-analysis, reporting the critical effect size values can caution readers that the mere attainment of statistical significance by the average effect does not automatically ensure scientific or real-life relevance.

To provide more tangible illustrations of the practical applications of critical effect size values, let us examine two real-world instances drawn from published research. The first one features a small sample, and has been targeted in the series of replications by the Open Science Collaboration [@open2015estimating], while the second one presents with a very large sample and has already attracted attention in the scientific community. The first case pertains to "Study 5" conducted by @mccrea2008self. This experiment compared two groups based on their performance percentage in a math test following a preliminary practice session with feedback of failure. One group was exposed to failure-excusing, self-handicapping thoughts, while both groups underwent the math test afterward. With a modest sample size of 28 participants split into two groups (13 and 15 participants, respectively), a directional one-tailed t-test was conducted on the percentage of correct answers to compare the groups. The resulting effect was statistically significant, $t(26) = 1.87, p (one-tailed) < .05$. Based on the reported data, Cohen's $d = 0.736$, indicating a considerable effect size. However, the critical value for Cohen's $d$ in this context was $0.645$, suggesting that, given this sample size, statistical significance is achieved only for estimated effects of substantial magnitude. Subsequent replication efforts [@open2015estimating] reaffirmed the significance of the effect, and used a somewhat larger sample ($n = 61$), $t(59)=2.325$. The estimated effect size was somehow lower than that originally reported, however, $d = 0.605$ (here, critical d = 0.428). The second scenario involves the study by @kramer2014experimental, which explored the impact of emotional content on Facebook users' experiences. With a notably large sample size of $n = 689'003$, the researchers observed a significant effect worsened emotional states when positive posts were reduced, $t(310,044) = −5.63$, $p < 0.001$ (non-directional test). Based on this and other results, the authors conclude that emotions expressed by other users on Facebook influence our own emotions. Despite the statistical significance, however, the critical values for Cohen's $d$ in this case are $-0.006$ and $+0.006$, emphasizing the importance of considering effect size beyond mere significance. This prompts reflection on whether an effect of such magnitude on such a large sample (actual $d = 0.02$) is truly meaningful in individuals' lives.
